---
title: "Practice - Vectors, Factors, Functions"
output: html_document
---

Vectors are objects that contain one or more elements. The elements of a vector should all be of the same type (mode) - numerical, logical, character, etc.

There are a few different ways of generating vectors of numbers. Let's try those.

## Using colon (:)

```{r}
1:10
```

```{r}
100:95
```
```{r}
-10:10
```
```{r}
-10:10
```

## Using `seq()`

```{r}
seq(0, 10)
```
```{r}
seq(0, 10, 2)
```
Note that the usual arithmetic operators work fine on vectors. This is one of the strengths of R.

```{r}
numbers <- seq(0, 10)
numbers + 1
```
Try other operators on the variable `numbers`


## Using `rep()`

This is for repeating the same item multiple times. `rep()` can be used for non-numeric vectors as well.


```{r}
rep(1.5, 10)
```
# 1

Vectors can be concatenated (joined) by using the `c()` function. Join `v1`, `v2`, and `v3` below to create a single vector.

```{r}
v1 <- seq(0, 0.9, 0.1)
v2 <-  1:9
v3 <- 1:10 * 10



```
# 2

What is the sum of all the elements in the combined vector in the above question?

```{r}

```

# 3

How do you calculate the average (mean) of a vector of numbers? One way is to compute the sum and divide by the number of elements in the vector. The number of elements is given by the `length()` function. There is also the `mean()` function that directly does this. Try both the ways on a vector.

```{r}

```

# 4

What is the sum of the first 100 integers? You may recall that the formula n(n+1)/2 gives this. Use the formula. Now, can you calculate it without using the formula? 

Hint: How do you generate a vector of numbers?

```{r}

```

# 5

Which is bigger - the sum of the first 100 odd numbers or the sum of the first 100 even numbers?

```{r}

```

# 6

Suppose we have a list of BMI values of several individuals. We want to generate a new vector with the BMI level information instead of the numerical values. We write a function to do this conversion.

```{r}
bmis <- c(15.22, 17.89, 29.91, 24.10, 34.98, 29.08, 19.32, 16.86, 26.65, 28.54, 25.11, 25.08, 18.36, 20.63, 16.29, 30.27, 24.92, 21.01, 24.98, 28.36, 21.50, 22.10, 26.29, 23.05, 27.58, 19.59, 21.54, 21.55, 16.85, 21.87, 26.63, 18.08, 25.05, 26.28, 21.98, 18.31, 22.04, 18.16, 25.07, 19.19, 30.19, 24.61, 21.18)

bmi_to_level <- function(bmi) {
  if(bmi < 18.5) {
    "Underweight"
  } else if(bmi < 25) {
    "Normal"
  } else if(bmi < 30) {
    "Overweight"
  } else {
    "Obese"
  }
}
```

But note that the function converts one single BMI value, not the whole vector. How do we convert the vector of numbers to a vector of levels? Of course, we can use a loop. Go over every single item in the BMI list and convert it.

```{r}
bmi_level <- c() # empty initial vector - this will be filled in inside the loop

for(i in 1:length(bmis)) {
  ## Do the correct assignment here.
  ## ith element of the bmis vector should be passed to the function
  bmi_level[i] <- 0
}

bmi_level
```
When you work with vectors and lists, it is more common in R to avoid using such loops and use the `apply()` family of functions. `sapply()` would be appropriate here. Check if the below chunk gives you the same result as the loop.

```{r}
bmi_level_apply <- sapply(bmis, bmi_to_level)
bmi_level_apply
```

If you notice, the level values are enclosed in quotes. This means they are strings. We can convert them to factors.

```{r}
bmi_level_apply <- factor(bmi_level_apply)
bmi_level_apply
```

Check the difference in the printed results.
