---
title: "Practice - Data Frame"
output: html_document
---

Data frames can be thought of as tabular data arranged in columns and rows. Each column contains a single type of information. In other words, each column is equivalent to a vector.

Let's organise some data from Census 2011 into a data frame. In practice, instead of creating it like this, we will be loading the data from a file (we will see this later).

```{r}
state_data <- data.frame(state = c("RAJASTHAN", "MADHYA PRADESH", "MAHARASHTRA",
                                   "ANDHRA PRADESH", "UTTAR PRADESH", "JAMMU & KASHMIR",
                                   "GUJARAT", "KARNATAKA", "ODISHA"),
                         households = c(12711146, 15093256, 24421519, 21022588,
                                        33448035, 2119718, 12248428, 13357027, 9637820),
                         persons = c(68548437, 72626809, 112374333, 84580777, 199812341,
                                     12541302, 60439692, 61095297, 41974218),
                         male = c(35550997, 37612306, 58243056, 42442146, 104480510,
                                  6640662, 31491260, 30966657, 21212136),
                         female = c(32997440, 35014503, 54131277, 42138631, 95331831,
                                    5900640, 28948432, 30128640, 20762082),
                         area = c(342239.00, 308252.00, 307713.00, 275045.00, 240928.00,
                                  222236.00, 196244.00, 191791.00, 155707.00))

```

The `head()` function shows you the top rows of the data frame.

```{r}
head(state_data)
```

Another way to print a data frame is using the `kable()` function in the knitr package.

```{r}
library(knitr)
kable(state_data)
```

You can also view the data frame by opening the variable directly from the environment section in Rstudio.


# 1

By using the `$` operator, each of the columns of the data frame may be accessed as separate vectors. E.g. `state_data$households`. Use this method to check the correlation between the population of a state and its area. You can use the `cor` function for this.


```{r}

```

# 2

Use vector division to calculate the sex ratio of these states.

```{r}

```

