---
title: "Practice - Data Wrangling"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
---

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, error = FALSE, message = FALSE)
```

# Census UA Data

This is an example using the Urban Agglomerations (Class-1 cities) population data from the Census of India 2011. We look at the population growth in these cities and the changes in sex ratio over the decades.

In this example, we use a cleaned version of the data. The data cleaning steps are shared as a separate file. 

```{r}
library(tidyverse)
data <- read_csv("../data/ua_populations.csv")
head(data)
```

# Sex ratio

We calculate sex ratio as the number of females per 1000 males.

## Overall

We can aggregate the population in the Class-1 Cities for each Census year and look at the change in sex ratio.

```{r}
data %>% 
  group_by(year) %>% 
  summarise(
    total_female = sum(pop_female, na.rm = TRUE),
    total_male = sum(pop_male, na.rm = TRUE)
  ) %>% 
  mutate(sex_ratio = round(total_female / total_male * 1000, 2))
```

We can clearly see an increasing trend overall from 1961 to 2011. Now let's focus on individual cities.

## Per City

```{r}
sr_data <- data %>% 
  mutate(sex_ratio = pop_female / pop_male * 1000) %>% 
  mutate(sex_ratio = round(sex_ratio, 2)) %>% 
  select(ua_no, ua, year, sex_ratio)

sr_data
```

Each city has a time series, making the data a bit difficult to consume. How do we look for patterns in this? First let's look at the latest situation.

### 2011 sex ratios

Filter the rows for the year 2011 and arrange the data by `sex_ratio`. Which cities have the largest sex ratios? Which ones have the smallest?

```{r}
sr_data %>% 
  filter(year == 2011) %>% 
  arrange(sex_ratio)
```

### Sex ratio change: 1991 to 2011

Many cities underwent a demographic transition over the decades. Several factors including migration played a role. How do we quantify this change? One simple way may be to take the difference in sex ratios across a couple of decades. Let us take 1991 as the base year and calcualte the change by 2011.

We will first need to change the shape of the data from long to wide. `pivot_wider()` is the function for it.

```{r}
sr_diff <- sr_data %>% 
  filter(year == 2011 | year == 1991) %>% 
  pivot_wider(names_from = year, values_from = sex_ratio)

sr_diff
```

Now we can calculate the difference across these years by just subtracting these two columns. Note that we need to use backticks (`) around the column names as they start with a number. Use arrange to order to check which cities saw the biggest changes in the positive and the negative direction.

```{r}
sr_diff %>% 
  mutate(sr_change = `2011` - `1991`)
```

# Exercise: Population Growth 

**Practice:** Do a similar analysis on population growth from 1991 to 2011. Are there cities that did not have much growth in population over these years?

```{r}
## Your code here

```



