---
title: "Practice - ggplot basics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, error = FALSE, message = FALSE)
```

`ggplot` is the grammar of graphics plotting library for R. The concept is to make plots by creating layers on a base plot in a step by step manner

# Load mtcars data

Run this block of code to load ggplot package and mtcars data

```{r}
library(ggplot2)
library(dplyr)
data(mtcars)
head(mtcars)
summary(mtcars)
```

Some data preparation now. We want the names of the cars in the data frame. They are now the row names, not a separate column. We can add the names as a column using `mutate`. We also change the number of cylinders to a factor variable for making the plots better.

```{r}
mtcars <- mtcars %>% 
  mutate(car_name = rownames(.)) %>% 
  mutate(cyl = factor(cyl, ordered = TRUE))
```

# 1. Scatter plot

`ggplot()` function creates an empty layer

```{r}
ggplot(mtcars)                                                # creates an emply base plot
```

`geom_point()` adds points with x and y coordinates, creating a scatter plot 

Let us plot weight against mileage

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +   # aestheic mapping - x, y
  geom_point()                           # add the geometric point object 
```

What if we want to see the effect of one more variable, number of cylinders? We can add it as colour in the scatter plot. Ggplot automatically shows the legend.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +   # add additional aesthetic color
  geom_point()
```

Now let's add one more variable, the engine horse power as the size of the points.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl, size = hp)) +  # add additional aesthetic size
  geom_point()
```

Try this yourself: Use ggplot to plot mpg against hp as points. Use colors to highlight the different cyl 

```{r}


```


# 2. Bar plots

Let us plot the mileage of all the cars in the data as a bar graph. We first add the car names as a column (it is in fact row labels). Then, using geom_bar with stat = "identity", we create the plot.

```{r}
mtcars %>% 
  ggplot() +  
  geom_bar(aes(x = car_name, y = mpg), stat = "identity")  
```

The x axis labels are unreadable. An easy solution is to flip the x and y axes.

```{r}
myplot <- mtcars %>% 
  ggplot() +  
  geom_bar(aes(y = car_name, x = mpg), stat = "identity") 

myplot
```

You might want to arrange the bars in increasing or decreasing order of mileage for better readability. We can do that in the aesthetics of geom_bar by specifying `x = reorder(car_name, mpg)`. Try it below.

```{r}
mtcars %>%
  ggplot() +  
  geom_bar(aes(y = reorder(car_name, -mpg), x = mpg), stat = "identity") 

```

## Themes

Save your last plot into a variable `myplot`. Try the different ggplot themes as below.

```{r}
myplot + theme_classic()                                                    # try out these themes
myplot + theme_gray()
myplot + theme_linedraw()
```

# 3. Finishing touches

```{r}
myplot + 
  theme_minimal() +
  labs(x = "Name of the car", y = "Miles per gallon") + # axis labels
  ggtitle("Cars and their mileage")   # plot title
```


We have only scratched the surface of what is possible with ggplot. If you would like to continue playing around with such example data sets and practice more ggplot features, this book chapter (freely available online) is a great resource: https://r4ds.had.co.nz/data-visualisation.html





