---
title: "Reproducible Data Analysis using R"
subtitle: 'Part 3: ggplot'
output:
  html_notebook:
    toc: yes
    toc_float:
      collapsed: no
  html_document:
    toc: yes
    toc_float:
      collapsed: no
    self_contained: yes
---

## Grammar of Graphics and Ggplot2

An influential book [Grammar of Graphics](https://www.cs.uic.edu/~wilkinson/TheGrammarOfGraphics/GOG.html) by Leland Wilkinson lays out a theoretical framework to systematically construct quantitative graphics. Hadley Wickham extended this and related ideas and implemented them in the ggplot2 package. If you are interested in more details, [this paper](http://vita.had.co.nz/papers/layered-grammar.pdf) would be a great starting point.  

At a very high level, this is what you would do with ggplot:

* create a plot with the `ggplot()` call
* add layers using various `geom_*()` or `stat_*()` calls. Some properties (e.g. position) of these layers can be set in these calls.
* specify "aesthetics" using `aes()` for the full plot or per-layer. This deals with selecting variables to plot, colours, sizes, shapes of the plot elements etc. Aesthetics can be "mapped" to variables in the data.
* optionally specify faceting
* optionally adjust the scale, the axes and the legend
* optionally customise the theme

**Resources:**

For a quick overview of ggplot, see https://r4ds.had.co.nz/data-visualisation.html.  

To learn and practice more, the Rmd files here are great: https://github.com/hadley/ggplot2-book

---

## Plotting the Urban Agglomerations

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, error = FALSE, message = FALSE)
```

We load the data saved by the tidy.Rmd file (part-2 of the tutorial).

```{r}
library(tidyverse)

data <- read_csv("data/ua_populations.csv")
sex_ratio_change <- read_csv("data/sex_ratio_change.csv")
```

Create a tidy population data set

```{r}
pop_data <- data %>% 
  rename(total = population, male = pop_male, female = pop_female) %>% 
  select(-c(area, pop_change, pop_change_percent)) %>% 
  gather(sex, population, total, male, female)

pop_data
```

## Male and Female Population Distributions

```{r}
pop_data %>% 
  filter(sex != "total" & year == 1961) %>% 
  ggplot(aes(x = log(population), fill = sex, alpha = 0.5)) +
    geom_density(position = "identity")
    
```

<br/>
Try changing `geom_density()` to `geom_histogram()`. Also try the `position = "dodge"` option. You could also add `facet_wrap()` to get the distributions for the years separately.

## Total Population Distributions over Time

```{r}
library(ggridges)

pop_data %>% 
  filter(sex == "total") %>% 
  ggplot(aes(x = log10(population), 
             y = factor(year, levels = rev(levels(factor(year)))))) +
    stat_density_ridges() 
```

<br/>`ggridges` is a ggplot extension. There are many such useful extensions, and you can see the list [here](https://www.ggplot2-exts.org/gallery/).  

## Line graphs - change over time


#### Plotting for one city - total population over time

```{r}
pop_data %>%
  filter(ua == "Vapi ^" & sex == "total") %>% 
  ggplot(aes(x = year, y = population)) +
    geom_line()
```

#### Facet wrap

```{r fig.width = 10, fig.height=6}
gap_increasing <- sex_ratio_change %>% 
  top_n(20, sex_ratio_slope)

gap_decreasing <- sex_ratio_change %>% 
  top_n(-20, sex_ratio_slope)

p <- pop_data %>%
  semi_join(gap_decreasing, by = "ua_no") %>% 
  filter(sex == "total") %>% 
  ggplot(aes(x = year, y = population)) +
    geom_line() +
    facet_wrap(~ua, scales = "free_y")

p + theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

#### Multiple cities - male vs. female

**Disparity closing**

```{r fig.width = 10, fig.height=8}
pop_data %>%
  semi_join(gap_decreasing, by = "ua_no") %>% 
  filter(sex != "total") %>% 
  ggplot() +
    geom_line(aes(x = year, y = population, colour = sex)) + 
    facet_wrap(~ua, scales = "free_y", ncol = 4, strip.position = "bottom") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

**Disparity increasing**

```{r fig.width = 10, fig.height=8}
pop_data %>%
  semi_join(gap_increasing, by = "ua_no") %>% 
  filter(sex != "total") %>% 
  ggplot() +
    geom_line(aes(x = year, y = population, colour = sex)) + 
    facet_wrap(~ua, scales = "free_y", ncol = 4, strip.position = "bottom") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

## 2011 population: bar plot

```{r}
pop_data %>% 
  filter(year == 2011 & sex == "total") %>% 
  arrange(desc(population)) %>% 
  top_n(30) %>% 
  ggplot() +
    geom_bar(aes(x = reorder(factor(ua), population), y = population), stat = "identity") +
    coord_flip() +
    xlab("City") +
    scale_y_continuous(label = scales::comma)
```

## Scatter plot

```{r fig.width = 10, fig.height=8}

cities_scatter <- data %>% 
  filter(year == 2011) %>% 
  mutate(
    sex_ratio = pop_male / pop_female,
    pop_density = population / area     
  ) %>%
  ggplot(aes(x = log10(population), y = pop_density, label = ua)) +
    geom_point(aes(size = log10(population),
                   colour = sex_ratio,
                   alpha = 0.4)) +
    scale_color_gradient(low = "green", high = "red") +
    scale_y_reverse()

cities_scatter
```

#### With labels

```{r fig.width = 8, fig.height=10}
library(ggrepel)

cities_scatter_labelled <- cities_scatter +
  geom_label_repel(aes(label = ifelse(sex_ratio > quantile(sex_ratio, 0.98, na.rm = TRUE) |
                                        sex_ratio < quantile(sex_ratio, 0.02, na.rm = TRUE) |
                                        pop_density > quantile(pop_density, 0.98, na.rm = TRUE) |
                                        pop_density < quantile(pop_density, 0.02, na.rm = TRUE), 
                                      ua, '')))

cities_scatter_labelled
```

## Interactive Scatter plot using plotly

Plotly is a library to create interactive visualisation. They have an API that follows the tidyverse philosophy. [This book](https://plotly-r.com/index.html) is a great resource to get started if you are interested. Plotly also provides a magic API to bring static ggplot plots to life. Here we will use only that.  


```{r fig.width = 8, fig.height=6}
plotly::ggplotly(cities_scatter)
```

## Scatter Plot Animation

Let's try something fancy!  

Ever felt like doing a [Hans Rosling](https://en.wikipedia.org/wiki/Hans_Rosling)? R lets you do some of that as well! In case you don't know what I am talking about, whatch [this video](https://www.ted.com/talks/hans_rosling_the_good_news_of_the_decade) (seek to 11:00 if you're in a hurry).  


```{r}
library(gganimate)

data %>% 
    semi_join(mega_cities, by = "ua_no") %>% 
    mutate(
      sex_ratio = pop_male / pop_female,
      pop_density = population / area     
    ) %>%
    ggplot(aes(x = sex_ratio, y = pop_density, label = ua)) +
      geom_point(aes(size = sqrt(population) / 300,
                     colour = ua,
                     alpha = 0.5)) +
      xlim(.8, 1.8) +
      ylim(25000, 0) +
      guides(size = FALSE, alpha = FALSE) + 
      scale_size_identity() +
      labs(title = 'Year: {frame_time}', x = 'Sex Ratio', y = 'Population Density') +
      transition_time(as.integer(year)) +
      ease_aes('linear')
```

## R - Links and Resources

* **BRUG**: [Bengaluru R UseR Group](https://www.meetup.com/en-AU/BengaluRu-use-R-gRoup/)
* [R Ladies](https://rladies.org/)
* [UseR! Conference](http://www.user2019.fr/)
* [R Bloggers](https://www.r-bloggers.com/)
* [The R Journal](https://journal.r-project.org/)





